\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-14 analysis/paper/paper.tex

\begin{CodeInput}
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
R+   family = "glmnet",
R+   time = "d.time", event = "death",
R+   formula_time = ~ log(d.time), alpha = 1,
R+   ratio = 10, standardize = TRUE,
R+   penalty.factor = c(0, rep(1, ncol(x)))
R+ )
\end{CodeInput}
\end{CodeChunk}

Next, we can use the object \code{cbFull} to estimate the absolute risk
curve on the test data. We first create a matrix \code{newx} containing
the relevant variables. Next, using \code{times}, we specify the time
points at which the cumulative incidence will be computed. Finally,
using the parameter \code{s = "lambda.1se"}, we specify at which value
of the tuning parameter we want to retrieve the coefficient estimates.

@@

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newdata parameter
R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
R+                   data = test)[, -c(1)]
R> times <- sort(unique(test$d.time))
R> 
R> abCbFull <- casebase::absoluteRisk(cbFull,
R+   time = times,
R+   newdata = newx,
R+   s = "lambda.1se",
R+   method = "numerical"
R+ )
\end{CodeInput}
\end{CodeChunk}

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of coefficients with values greater than 0 are displayed in
\ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression to have
larger estimated coefficients. This is expected, as there are no
restrictions on the size of coefficients. The smallest coefficient
estimates are found in penalized case-base model, aside from cancer
state (cano).

@@

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized case-base model compared to both Cox regression models. The
differences, however, are minimal. All three outperform the Kaplan-Meier
estimator.

@@

\caption{\label{fig:cs3FinalBrier} Comparison of Cox regression, penalized Cox regression, penalized case-base estimation, and Kaplan-Meier. (A)  Probabilities of death as it changes over follow-up time. (B) Brier scores as they change over time, where a lower score corresponds to better performace. Penalized case-base estimation produced the lowest results over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

@@

This report was generated on 2020-08-31 10:05:10 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.4             glue_1.4.1               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.2              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.1               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.1.0          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.0.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_1.13.3    base64enc_0.1-3    
#>  [7] fs_1.4.1            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.0      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.8     assertthat_0.2.1   
#> [25] cli_2.0.2           acepack_1.4.1       htmltools_0.5.0    
#> [28] quantreg_5.55       tools_4.0.2         gtable_0.3.0       
#> [31] Rcpp_1.0.5          cellranger_1.1.0    vctrs_0.3.2        
#> [34] nlme_3.1-148        iterators_1.0.12    xfun_0.15          
#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.17   
#> [40] muhaz_1.2.6.1       MASS_7.3-51.6       zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.14        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_5.1-4           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     evaluate_0.14      
#> [61] lattice_0.20-41     labeling_0.3        htmlwidgets_1.5.1  
#> [64] cmprsk_2.2-10       tidyselect_1.1.0    deSolve_1.28       
#> [67] plyr_1.8.6          R6_2.4.1            generics_0.0.2     
#> [70] Hmisc_4.4-0         multcomp_1.4-13     DBI_1.1.0          
#> [73] withr_2.2.0         pillar_1.4.6        haven_2.3.1        
#> [76] foreign_0.8-80      mgcv_1.8-31         nnet_7.3-14        
#> [79] mstate_0.2.12       modelr_0.1.8        crayon_1.3.4       
#> [82] rmarkdown_2.3       jpeg_0.1-8.1        grid_4.0.2         
#> [85] readxl_1.3.1        blob_1.2.1          reprex_0.3.0       
#> [88] digest_0.6.25       webshot_0.5.2       numDeriv_2016.8-1.1
#> [91] stats4_4.0.2        munsell_0.5.0       viridisLite_0.3.0  
#> [94] quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}

>>>>

\begin{CodeInput}
R> x <- model.matrix(death ~ . - d.time - aps - sps, 
+                   data = train)[, -c(1)] # Remove intercept
R> y <- data.matrix(subset(train, select = c(d.time, death)))
R> 
R> # Regularized logistic regression to estimate hazard
R> cbFull <- casebase::fitSmoothHazard.fit(x, y,
+   family = "glmnet",
+   time = "d.time", event = "death",
+   formula_time = ~ log(d.time), alpha = 1,
+   ratio = 10, standardize = TRUE,
+   penalty.factor = c(0, rep(1, ncol(x)))
+ )
\end{CodeInput}
\end{CodeChunk}

Next, we can use the object \code{cbFull} to estimate the absolute risk
curve on the test data. We first create a matrix \code{newx} containing
the relevant variables. Using \code{times}, we specify the time points
at which the cumulative incidence will be computed. Finally, using the
parameter \code{s = "lambda.1se"}, we specify at which value of the
tuning parameter we want to retrieve the coefficient estimates.

@@

\begin{CodeInput}
R> # Estimating the absolute risk curve using the newdata parameter
R> newx <- model.matrix(death ~ . - d.time - aps - sps, 
+                   data = test)[, -c(1)]
R> times <- sort(unique(test$d.time))
R> #times<-seq(from=min(times),to=max(times),length.out = 100)
R> 
R> abCbFull <- casebase::absoluteRisk(cbFull,
+   time = times,
+   newdata = newx,
+   s = "lambda.1se",
+   method = "numerical"
+ )
\end{CodeInput}
\end{CodeChunk}

@@

Our first comparison plot examines the covariates that were selected
between all three hazard functions. Note that Cox regression without
penalization selected more covariates than the other two, but only the
intersection of coefficients with values greater than 0 are displayed in
Figure \ref{fig:cs3lolliPlot}. A visible tendency is for Cox regression
to have larger estimated coefficients. This is expected, as there are no
restrictions on the size of coefficients. The smallest coefficient
estimates are found in penalized logistic model, aside from cancer state
(cano).

@@

\caption{\label{fig:cs3FinalBrier} (A) Four curves comparing the probabilities of death as it changes over follow-up time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. All follow-up times seen in the new data were used to generate the penalized casebase and K-M curves. The Cox-based curves, estimated using the survival package, appear to be smooth because all 1703 follow-up times seen while fitting are used. Using user specified times is not permitted in the survival package. (B) Four curves demonstrating the Brier scores as they change over time between Cox regression, penalized Cox regression, penalized logistic regression, and Kaplan-Meier. Lower is better. Penalized logistic regression produced the lowest Brier scores over follow-up time.}\label{fig:riskregressionBrier}
\end{figure}
\end{CodeChunk}

These values do not inform us of model performance. To do so, we take
our hazard functions and use them to predict the absolute risk of
everyone in our test set, over time. The probabilities over time for
each individual are averaged, resulting in the absolute risk curves in
Figure \ref{fig:cs3FinalBrier} (A). There is separation between the
curves, but it is minimal. To better demonstrate the differences in
performance between the models, we compare Brier scores over time. In
Figure \ref{fig:cs3FinalBrier} (B), the Brier score is lower in our
penalized logistic regression model compared to both Cox regression
models. The differences, however, are minimal. All three outperform the
Kaplan-Meier model.

@@

This report was generated on 2020-09-14 12:43:13 using the following
computational environment and dependencies:

@@

\begin{CodeOutput}
#> R version 4.0.2 (2020-06-22)
#> Platform: x86_64-apple-darwin17.0 (64-bit)
#> Running under: macOS Catalina 10.15.6
#> 
#> Matrix products: default
#> BLAS:   /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRblas.dylib
#> LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib
#> 
#> locale:
#> [1] en_CA.UTF-8/en_CA.UTF-8/en_CA.UTF-8/C/en_CA.UTF-8/en_CA.UTF-8
#> 
#> attached base packages:
#> [1] splines   stats     graphics  grDevices utils     datasets  methods  
#> [8] base     
#> 
#> other attached packages:
#>  [1] broom_0.7.0               dotwhisker_0.5.0         
#>  [3] timereg_1.9.7             glue_1.4.2               
#>  [5] scales_1.1.1              forcats_0.5.0            
#>  [7] stringr_1.4.0             dplyr_1.0.2              
#>  [9] purrr_0.3.4               readr_1.3.1              
#> [11] tidyr_1.1.2               ggplot2_3.3.2            
#> [13] tidyverse_1.3.0           lubridate_1.7.9          
#> [15] visreg_2.7.0              tibble_3.0.3             
#> [17] riskRegression_2020.02.05 prodlim_2019.11.13       
#> [19] flexsurv_1.1.1            survival_3.2-3           
#> [21] pracma_2.2.9              magrittr_1.5             
#> [23] kableExtra_1.2.1          glmnet_4.0-2             
#> [25] Matrix_1.2-18             data.table_1.13.0        
#> [27] cowplot_1.1.0             colorspace_1.4-1         
#> [29] casebase_0.9.0.9000      
#> 
#> loaded via a namespace (and not attached):
#>  [1] TH.data_1.0-10      VGAM_1.1-3          ellipsis_0.3.1     
#>  [4] ggstance_0.3.4      htmlTable_2.0.1     base64enc_0.1-3    
#>  [7] fs_1.5.0            rstudioapi_0.11     farver_2.0.3       
#> [10] MatrixModels_0.4-1  fansi_0.4.1         mvtnorm_1.1-1      
#> [13] xml2_1.3.2          codetools_0.2-16    knitr_1.29         
#> [16] Formula_1.2-3       jsonlite_1.7.1      cluster_2.1.0      
#> [19] dbplyr_1.4.4        png_0.1-7           compiler_4.0.2     
#> [22] httr_1.4.2          backports_1.1.9     assertthat_0.2.1   
#> [25] cli_2.0.2           htmltools_0.5.0     quantreg_5.67      
#> [28] tools_4.0.2         gtable_0.3.0        Rcpp_1.0.5         
#> [31] cellranger_1.1.0    vctrs_0.3.4         nlme_3.1-149       
#> [34] conquer_1.0.2       iterators_1.0.12    xfun_0.17          
#> [37] rvest_0.3.6         lifecycle_0.2.0     polspline_1.1.19   
#> [40] muhaz_1.2.6.1       MASS_7.3-53         zoo_1.8-8          
#> [43] hms_0.5.3           sandwich_2.5-1      SparseM_1.78       
#> [46] RColorBrewer_1.1-2  rticles_0.15        yaml_2.2.1         
#> [49] gridExtra_2.3       rms_6.0-1           rpart_4.1-15       
#> [52] latticeExtra_0.6-29 stringi_1.4.6       foreach_1.5.0      
#> [55] checkmate_2.0.0     lava_1.6.7          shape_1.4.4        
#> [58] rlang_0.4.7         pkgconfig_2.0.3     matrixStats_0.56.0 
#> [61] evaluate_0.14       lattice_0.20-41     labeling_0.3       
#> [64] htmlwidgets_1.5.1   cmprsk_2.2-10       tidyselect_1.1.0   
#> [67] deSolve_1.28        plyr_1.8.6          R6_2.4.1           
#> [70] generics_0.0.2      Hmisc_4.4-1         multcomp_1.4-13    
#> [73] DBI_1.1.0           withr_2.2.0         pillar_1.4.6       
#> [76] haven_2.3.1         foreign_0.8-80      mgcv_1.8-33        
#> [79] nnet_7.3-14         mstate_0.2.12       modelr_0.1.8       
#> [82] crayon_1.3.4        rmarkdown_2.3       jpeg_0.1-8.1       
#> [85] grid_4.0.2          readxl_1.3.1        blob_1.2.1         
#> [88] reprex_0.3.0        digest_0.6.25       webshot_0.5.2      
#> [91] numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
#> [94] viridisLite_0.3.0   quadprog_1.5-8
\end{CodeOutput}
\end{CodeChunk}
